---
title: "Use case for the Multivariate Source Entropy Triangle"
author: "Francisco J. Valverde-Albacete & Carmen Peláez Moreno"
date: "14 de diciembre de 2015"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Environment construction

```{r}
#library(ggtern)   # Excellent package for ternary diagrams in the gg tradition
#library(entropy)  # To work out the appropriate coordinates.
library(dplyr)     # That infamous Mr. Wickham!
library(infotheo)  # The functionality provided by this has to be rerouted through entropies
library(entropies) # This package
library(ggtern)    # Ternary diagrams on ggplot
library(vcd)       # Categorical benchmarks
library(mlbench)   # ml benchmarkss
```


# An example of multivariate use

## Datasets available

The inventory of databases to be explored:

```{r}
# the inventory of databases you can access
#library(datasets)
name <- c("Ionosphere", "iris", "Glass", "Arthritis", "BreastCancer") # 
className <- c("Class","Species", "Type", "Improved", "Class")  # Name of class attribute
classVar <- c(35, 5, 10, 5, 11)   # ordinal of the class attribute
other <- c("", "", "", "", "ID", "Id") # Other attributes to dispose of: mainly identifiers.
K <- c(2, 3, 7, 3, 2)  # No. of classes
datasets <- data.frame(name,classVar,className,K, stringsAsFactors=FALSE)

# To select a dataset by name
# Caveat: you have to ensure that the containing package has been attached
evalDataset <- function(dsName){
    dsName <- as.character(dsName)
    switch(dsName,
        "iris" =         {data(iris); iris},
        "Ionosphere" =   {data(Ionosphere); Ionosphere},
        "Glass" =        {data(Glass); Glass},
        "Arthritis" =    {data(Arthritis); Arthritis},
        "BreastCancer" = {data(BreastCancer); BreastCancer}
        )
}
# #To select the #of column of the classc
# whichClass <- function(ds, className){which(colnames(evalDataset(ds))==className)}
# #whichNumVar <-  function(r){whichClass(evalDataset(r$name), r$className)}
# cardinalClass <- function(ds, className){
#     length(unique(evalDataset(ds)[,className]))
# }
# classVar <-  mapply(whichClass, datasets$name, datasets$className)
# K <- mapply(cardinalClass, datasets$name, classVar)
# library(dplyr)
# datasets <- data.frame(name,className, classVar, K)
datasets
# Let's print it on latex
library(xtable)
thisLatex <- xtable(datasets, caption="Some datasets considered in this study")
align(thisLatex) <- xalign(thisLatex)
thisLatex
```

## Obtaining the entropies

Obtain the entropies and some other data for plotting from all datasets.

```{r}
# Analyze both with Classes and without
withClasses <- c(TRUE, FALSE)
#  "name"       "H_Uxi"      "H_Pxi"      "DeltaH_Pxi" "VI_Pxi"     "M_Pxi" 
# "withClass", "dsName", "isClass"
edf <- data.frame()
for(dsName in unique(datasets$name)){
    dsRecord <-  filter(datasets, name == dsName)
    ds <- evalDataset(dsName)
    for(withClass in withClasses){
        if (withClass){
            print(sprintf("Analyzing dataset with class label: %s", dsName))
        }else {
            print(sprintf("Analyzing dataset without class label: %s", dsName))
            # as per: 
            # http://stackoverflow.com/questions/5234117/how-to-drop-columns-by-name-in-a-data-frame
            # Don't EVER use subset in PROGRAMS!
            #ds <- subset(ds, subset=1:nrow(ds), select=dsRecord$className, drop=TRUE)
            ds <- ds[, !colnames(ds) == dsRecord$className] #fastest in bechnmark at bot. of url
        }
        edf <- rbind(edf,
                     sentropies(ds) %>% 
#                     sentropies(ds, nbins=ceiling(nrow(ds)^(1/3))) %>%
                         mutate(withClass = withClass,
                                isClass = (as.character(dsRecord$className) ==
                                                as.character(name)),
                                dsName = dsName
                         )
        )
    }
}
str(edf)
```


## Plotting the aggregate data

Now we print the total entropy balance in the triangle

```{r}
excised <- c("name", "withClass", "isClass", "dsName") #Or any other that we add
totalEdf <-  edf %>% 
            group_by(dsName, withClass) %>%
            summarise(H_Ux = sum(H_Uxi),
                    H_Px = sum(H_Pxi),
                    DeltaH_Px = sum(DeltaH_Pxi),
                    VI_Px = sum(VI_Pxi),
                    M_Px = sum(M_Pxi))
totalEdf
#Use the dsName for shape and the withClass for fill => use only fillable glyphs.
smet <-  ggmetern(totalEdf) + geom_point(mapping=aes(shape=dsName, color=withClass), size=3)
    #+ ggtitle("Aggregate entropy by dataset")
smet +  scale_colour_manual(values = c("black", "red")) +
    labs(shape="dataset", colour="Using class")
```

## Plotting the multisplit data

We choose some of the interesting datasets from the diagram above to investigate:


## Exploring individual datasets

```{r}
#thisDsName <- "Ionosphere" # CAVEAT! Not enough different glyphs!!!
#thisDsName <- "iris"
#thisDsName <- "Glass"
thisDsName <- "Arthritis"
thisDsName <- "BreastCancer"
# negatively subsetting recipe from Stack Overflow
thisEdf <-  rbind(edf %>% filter(dsName == thisDsName & name != "ALL") %>% 
                      select(-starts_with("isClass")), #%>% 
                      #filter(name != "ALL"),
                  totalEdf %>% filter(dsName == thisDsName) %>% mutate(name = "@AGGREGATE") %>%
                      select(name, #matches(".*"), 
                             H_Uxi = H_Ux, H_Pxi = H_Px, DeltaH_Pxi = DeltaH_Px,
                             VI_Pxi = VI_Px, M_Pxi = M_Px,
                             withClass, dsName)
                ) 
#edf1 <- entropies(df, nbins=ceiling(nrow(df)^(1/3)))
#edf1
thisEt <-  ggmetern(thisEdf) + geom_point(size=4, aes(shape=name, colour=withClass))
thisEt + scale_colour_manual(values = c("black", "red")) +
    scale_shape_manual(values=1:14) + 
    labs(shape="feature", colour="Using class")
```


# The absolute decomposition

This actually rather resembles a set of stacked pies as seen from above!

```{r}
```

```{r}
# In case the switch is ON for excluding the aggregate.
excludeAggregate <- TRUE
#excludeAggregate <- FALSE

analyzeWithClass <- FALSE
#analyzeWithClass <- TRUE
if (analyzeWithClass){
    smedf <- filter(thisEdf, withClass) %>% select(-withClass)
} else {
# For this once let's just use the entropy with no class variable
    smedf <- filter(thisEdf, !withClass) %>% select(-withClass) # source multivariate entropy data frame
}
#library(scales)
p <-  ggmebars(smedf, excludeAggregate,proportional=FALSE)
p + ylab("Source Multivariate Entropy") + xlab("Feature/Variable") #+ ae__
    # scale_y_continuous(trans=log2_trans(),
    #                 breaks=trans_breaks("log2", function(x) 2^x),
    #                 labels=trans_format("log2", math_format(2^.x)))
# 
```


The multisplit SMET can also offer the same information with a different encoding: using the fill color on the glyphs used to represent the features:

```{r}
ggmetern(filter(smedf, name != "@AGGREGATE")) +
    geom_point(size=4, aes(shape=name, colour=H_Uxi)) + scale_shape_manual(values=1:14) + 
    labs(shape="Feature", colour="$H_{U_{X_i}}") #+ theme(legend.position="bottom")
    #geom_point(size=4, aes(shape=name, colour=withClass))
```


However, turning the stacked bar graph into a pie chart is a bad idea, since the remaining information $VI_{P_{X_i}}$ is de-emphasized:

```{r}
 p + ylab("Source Multivariate Entropy") + xlab("Feature/Variable") + coord_polar()
```

# SO FAR SEEMS OK
## Facetting on different values


```{r}
selected <- c("iris", "Glass", "BreastCancer")
#selEdf <- edf %>% filter(dsName %in% selected)
thisSelEdf <-  rbind(edf %>% filter(dsName %in% selected) %>%select(-starts_with("isClass")),
                  totalEdf %>% filter(dsName %in% selected) %>% mutate(name = "AGGREGATE") %>%
                      select(name, #matches(".*"), 
                             H_Uxi = H_Ux, H_Pxi = H_Px, DeltaH_Pxi = DeltaH_Px,
                             VI_Pxi = VI_Px, M_Pxi = M_Px,
                             withClass, dsName)
                )
#edf1 <- entropies(df, nbins=ceiling(nrow(df)^(1/3)))
#edf1
thisSelEt <-  ggmetern(thisEdf, size=4, aes(shape=name)) #, colour=withClass))
thisSelEt + facet_grid(dsName ~ withClass) + 
    scale_colour_manual(values = c("black", "red")) +
    scale_shape_manual(values=1:14) + 
    labs(shape="feature") #s, colour="Using class")
```

# Use on different datasets

Get some categorical datasets from package "vcd"
```{r}
library(vcd)
library(mlbench)
pname <-  "vcd" # categorical variables
pname <- "mlbench"
ps  <- data(package=pname)     # get all the info for the data in the package
vcdData <- ps$results[,"Item"] # get the name of all the data 
#Now find the names of the datasets that are categorical!
vcdDataframes  <- vcdData[unlist(lapply(ps$results[,"Item"],
                                        function(x){is.data.frame(eval(as.name(x)))})
                                 )
                          ]
#list them
vcdDataframes
```

Do the same treatment on Arthritis:

```{r}
withClass <-  TRUE
withClass <-  FALSE
if (withClass){# Consider the class attribute
    df <- iris
    #df <- Arthritis
    #df <- BreastCancer[,-1] # Take away identifiers.
    df <- {data(Glass); Glass}
    #df <- {data(DNA); DNA}
}else{ #Do not consider the class attribute or e.g. identities.
    df <- iris[, -c(5)]
    #df <- Arthritis[,c(2:4)]
    #df <- BreastCancer[,-c(1,11)] # 1 is an identifier, 11 is the Class
    df <- Glass[,-c(10)] # 10 = Type and identifier of the type of Glass!
    #df <- DNA[,-c(181)] # V. 181 = Class
}
edf <- entropies(df, nbins=ceiling(nrow(df)^(1/3)))
# Number one is always the name of the features!
addTotalEntropies <- function(edf, label="@Total"){
  edf <-cbind(
    data.frame(name=c(as.character(edf[,1]), label)),
    rbind(edf[,-1], colSums(edf[,-1]))
    )
  return(edf)
}
edf <- addTotalEntropies(edf)
ep <-  ggmetern(edf) + labs(shape="Feature")
ep
```
