---
title: "Analysis of Confusion Matrices"
author: "F.J. Valverde-Albacete and C. Pelaez-Moreno"
date: "June, 21st, 2018"
output: html_document
vignette: > 
  %\VignetteIndexEntry{A first guide to use the CBET for confusion matrix evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
This script shows how to read a Channel Binary Entropy Triangle (CBET) using different confusion matrices from (Valverde-Albacete & Peláez-Moreno, 2010).


### Environment construction
```{r setup, include=FALSE}
library(tidyverse) # That (in)famous Mr. Wickham!;)
#library(ggtern)    # Ternary diagrams on ggplot
library(entropies) # This package. Depends heavily on "ggtern", entropy", "infotheory".
```

Some top level switches and options gathered in one place. 
```{r switches}
debugLevel <- 0 # Debug level 0-non-existent, 1-minimal, the greater the more verbose.
fancy <- TRUE  # set this for nicer on-screen visualization.
#fancy <- FALSE # Set this for either printing matter or more austere plots.
getPlot <- TRUE # Flag to obtain plots for publication. 
getPlot <- FALSE #Comment to get .jpeg files rather than plots of ets.
knitr::opts_chunk$set(comment=NA, fig.width=6, fig.height=4)
if (getPlot)
    knitr::opts_chunk$set(dev = 'pdf') # plots in pdf, better for publication
```

### Data

```{r, data}
mats <- list(
    a = c(15, 0, 0, 0, 15, 0, 5, 5, 20),
    b = c(16, 2, 1, 2, 16, 1, 2, 2, 18),
    c = c(1,0,1,0,1,1,4,4,48),
    d = c(15,0,0,0,18,0,0,0,27),
    e = c(1,0,0,0,2,0,0,0,57),
    f = c(0,0,0,0,0,0,5,5,50)
)
as.atable <- function(m){matrix(m, nrow=3,ncol=3)}
cm <- lapply(mats,as.atable)
cmNames <- names(mats)
cm
```

### Entropies

```{r, entropies}
etdf <- data.frame()
for (i in 1:length(cmNames)) {
    etdf <- rbind(etdf,
                  cbind(name=cmNames[[i]], jentropies(as.table(cm[[i]])))
    )
}
```

# Visualizing entropies in the CBET

```{r, visualization}
confusion_cbet <- 
    ggmetern(etdf %>% filter(type == "XY"), fancy) +
    geom_point(aes(shape=name), size=3, color="blue") +
    labs(shape="Confusion Matrix") +
    theme(legend.key=element_blank())
confusion_cbet
if (getPlot){
    dev.off()#Necessary to do the textual plot.
    ggsave(stringr::str_interp("confusion_matrices_CBET_PRL10_in_R.jpeg"),
           plot=confusion_cbet)
}
```

The analysis of the classifiers generating the matrices is the following:
1. That generating a is an excellent classifier, transferring a lot of information 

# Visualizing Entropies in the split CBET

```{r, split visualization}
confusion_split_cbet <- 
    ggmetern(etdf, fancy) %+% 
    geom_point(aes(shape=type), 
               size=3, 
               color="blue") +
    scale_shape_manual(values=c("X"=4, "Y"=1, "XY"=20)) +
    labs(shape="Split Confusion Matrix") +
    theme(legend.key=element_blank())
confusion_split_cbet <-  
    confusion_split_cbet + 
    geom_text(data=etdf %>% filter(type == "XY"), aes(label=name), 
              color="blue",size=4, vjust=2, hjust=1)
confusion_split_cbet
if (getPlot){
    dev.off()#Necessary to do the textual plot.
    ggsave(stringr::str_interp("confusion_matrices_split_CBET_PRL10_in_R.jpeg"),
           plot=confusion_split_cbet)
}

```

# Session information

```{r}
sessionInfo()
```
