---
title: "SMET Reading aid"
author: "Francisco J. Valverde-Albacete & Carmen Peláez Moreno"
date: "10 de mayo de 2016"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A guide to understanding the positions in the SMET}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignettes tries to present instances of features that have specific stochastic properties to improve the understanding and intution of what each corner of the Source Multivariate Entropy Triangle (SMET) means.
 

# Environment construction

```{r environment, echo=FALSE, results="hide"}
#library(ggtern)   # Excellent package for ternary diagrams in the gg tradition
#library(entropy)  # To work out the appropriate coordinates.
library(dplyr)     # That infamous Mr. Wickham!
library(tidyr)     # Tidying tall & wide dataframes
library(infotheo)  # The functionality provided by this has to be rerouted through entropies
library(entropies) # This package
library(ggtern)    # Ternary diagrams on ggplot
library(vcd)       # Categorical benchmarks
library(mlbench)   # ml benchmarkss
#knitr::opts_chunk$set(dev = 'pdf') # plots in pdf, better for publication
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment=NA, fig.width=6, fig.height=4)
```

# Conceptual triangle

This is the conceptual triangle for the AGGREGATE source multivariate entropy triangle:

```{r}
m <- 
    rbind(
    c("irredundant", 0.00, 0.00, 1.0, "irredundant"), 
    c("redundant", 0.0, 1.0, 0.0, "redundant"),
    c("deterministic", 1.0, 0.0, 0.0, "deterministic"),
    c("almost irredundant", 0.05, 0.05, 0.9, "irredundant"), 
    c("almost redundant", 0.05, 0.9, 0.05, "redundant"),
    c("almost deterministic", 0.9, 0.05, 0.05, "quasi-deterministic")
)
d <- data.frame(
    name = m[,1],
    DeltaH_Px=as.numeric(m[,2]),
    M_Px=as.numeric(m[,3]),
    VI_Px=as.numeric(m[,4]),
    point = m[,5]
) %>% mutate(extreme = DeltaH_Px == 1.0 | M_Px == 1.0 | VI_Px == 1.0)
#multiSplitEntropies  <-  splitESmetCoords <-  c("name", "DeltaH_Pxi", "M_Pxi", "VI_Pxi") 
#names(d) <-  c(multiSplitEntropies, "point", "extremePoint")
```

```{r}
# THe points at the vertices
conceptual <- ggmetern(filter(d, extreme), fancy=FALSE) + 
    geom_point(aes(fill=name, shape=name), size=3) +    
    #scale_fill_manual(values=c("red", "blue", "orange")) +
    scale_shape_manual(values=c(21,22,23)) + #, guide=FALSE) +
    #guides(fill=FALSE, shape=FALSE) 
    guides(fill=FALSE) + labs(shape="Source type") 
    #labs(shape="Feature type") + #guides(color=FALSE) #scale_color_manual(guide=FALSE)
    #guides(fill=FALSE,shape=FALSE) #scale_color_manual(guide=FALSE)

conceptual # just prints the extreme points.
#Now place the non-extreme points: be careful to rename them
nonExtremes <- filter(d, !extreme) %>%
    #rename(Delta=DeltaH_Pxi, M=M_Pxi, VI=VI_Pxi) %>%
    select(-point, -extreme)

conceptual + geom_point(data=nonExtremes, aes(shape=name), size=3)

# #conceptual +  geom_text(x=0.05, y=0.05, label="HELLO")
# x_coords <- d[4:6, "DeltaH_Pxi"]
# y_coords <- d[4:6, "M_Pxi"]
# labs <- d[4:6, "name"]
# conceptual + geom_text(x=x_coords, y=y_coords, label=labs)
# 
# 
# conceptual +  geom_text(x=c(0.2), label="HELLO")
# conceptual + annotate(geom="text", x=0.05, y=0.05, label="XYZ", color="red")

#the points in the middle of the angles and dise
conceptual <- conceptual + 
    geom_point(data=filter(d, !extreme), aes(shape=name), size=3) +
    #scale_shape_manual(values=rep(22,3), guide=FALSE) + 
    scale_shape_manual(values=rep(c(21,22,23),2)) + #, guide=FALSE) + 
    #scale_colour_manual(values=c("red", "blue", "orange")) +
    #guides(shape=guide_legend()) + #,fill=guide_colorbar()) + 
    labs(shape="Source type")
    #guides(shape=FALSE) +
    
    #labs(shape="Source type") +
    #geom_text(aes(label=point, hjust="inward", vjust="inward"))
    #geom_text(aes(label=point, hjust="inward", angle=45)) # Does NOT work!
    #annotate("segment", x=1950, xend=1980, )
    # geom_point(aes(shape = name), size=3) + 
    # scale_shape_manual(values=c(21, 22, 24)) +
    # #tern_limits(T=1.2, L=1.2, R=1.2) + # This afffect the contour DON'T USE IT!
    # scale_T_continuous(expand=c(0.1, 0.1)) +
    # scale_L_continuous(expand=c(0.1, 0.1)) +
    # scale_R_continuous(expand=c(0.1, 0.1)) +
    # theme_bw() +
    # theme_rotate(degrees=-60) + 
    # theme(
    #     #tern.axis.col.T="black",
    #     #tern.panel.background = element_rect(colour="white"), 
    #     tern.axis.line = element_line(colour="black"),
    #     tern.axis.ticks.minor = element_line(colour="black"),
    #     #tern.axis.ticks.length.minor = unit(5, units="points"),
    #     tern.panel.grid.major = element_line(colour="grey", linetype=3), #too coarse
    #     tern.axis.text.show = TRUE,
    #     tern.axis.arrow.show = TRUE
    #     )
conceptual
ggsave(filename="annotated_SMET_triangle.jpeg", plot=conceptual)
```


# Data generation

This describes a split triangle.

```{r}
m <- 240 # vector instances
nRandom <- 5     # Add a number of balanced, random features
dsRandom <- as.data.frame(matrix(rep(0, m*nRandom), nrow=m, ncol=nRandom)) 
# Add the random balanced vectors
rRandom <- 1:nRandom
set.seed(29)
for(f in rRandom){#generate random features: 
    dsRandom[,f] <- rbinom(m, 1, 0.5)
}
names(dsRandom) <- paste0("b",rRandom)
##########################################################################################
nUnbalanced <- 5 # Add a number of unbalanced, random features
dsUnbalanced <- as.data.frame(matrix(rep(0, m*nUnbalanced), nrow=m, ncol=nUnbalanced))
# Add the random unbalanced vectors.
rUnbalanced <-  1:nUnbalanced
for(f in rUnbalanced){#generate random features: 
    dsUnbalanced[,f] <- rbinom(m, 1, 0.99)
}
names(dsUnbalanced) <-  paste0("u", rUnbalanced)
##########################################################################################
nRedundant <- 5  # Add a number of very redundant features
dsRedundant <- as.data.frame(matrix(rep(0, m*nRedundant), nrow=m, ncol=nRedundant))
#n <- nRandom + nUnbalanced + nRedundant
#ds <- as.data.frame(matrix(rep(0, m+n), nrow=m, ncol=n))
# Add a number of very redundant vectors
rRedundant <- 1:nRedundant
v <- as.numeric(rbinom(m, 1, 0.5)) # They are also quite balanced
for(f in rRedundant){
    #dsRedundant[,f] <- jitter(v * f, amount=0.1*f)
    dsRedundant[,f] <- jitter(v, factor=0.5)
    #dsRedundant[,f] <- as.numeric(rbinom(m, 1, 0.5)) + 05
}
names(dsRedundant) <- paste0("r", rRedundant)

##########################################################################################
#names(ds) <- c(paste0("b",rRandom), paste0("u", rUnbalanced), paste0("r", rRedundant))
```


# Visualization

```{r}
###########################################################################
edsRandom <- sentropies(dsRandom) %>% mutate(type="irredundant")
            #(type=substr(name, 1,1)) %>% filter(type != "A")
ggmetern(edsRandom) + # now create the plot
    geom_point(aes(colour=name), size=3)
edsUnbalanced <- sentropies(dsUnbalanced) %>% mutate(type="quasi-deterministic")
    #mutate(type=substr(name, 1,1)) %>% filter(type != "A")
ggmetern(edsUnbalanced) + # now create the plot
    geom_point(aes(colour=name), size=3)
edsRedundant <- sentropies(dsRedundant) %>% mutate(type="redundant")
ggmetern(edsRedundant) + # now create the plot
    geom_point(aes(colour=name), size=3)
###########################################################################
# Put together al the different types of points
eds <- rbind(edsRandom, edsUnbalanced, edsRedundant) %>% filter(name != "ALL")
#summary(eds)
```

```{r}
# First add the densities
#et <- ggmetern(eds, fancy=FALSE) #+ # now create the plot
    # stat_density_tern(geom='polygon',
    #                      n         = 200,
    #                      aes(fill  = ..level..,
    #                          alpha = ..level..)) +    
    # scale_fill_gradient(low = "white",high = "black")  +
    # guides(fill = "none", alpha= "none")
    #guides(fill = "none", alpha = "none")
# then add the points on the original data
et <-  ggmetern(eds, fancy=FALSE) + 
    geom_point(aes(shape=type), size=3) +
    scale_shape_manual(values=c(21,22,23)) + #, guide=FALSE) + 
    scale_colour_manual(values=c("red", "blue", "orange")) +
    labs(shape="Feature type") + guides(color=FALSE) #scale_color_manual(guide=FALSE)
    #scale_fill_manual(values=c("red", "blue", "yellow"))
et
ggsave(filename="annotated_multisplit_SMET.jpeg", plot=et)
```

# Postscriptum

```{r ps}
sessionInfo()
```
