---
title: "Infodynamics of classification"
author: "Francisco J. Valverde-Albacete & Carmen Peláez Moreno"
date: "9 de mayo de 2016"
output: html_document
---
# Introduction

This vignette tries to demonstrate the use of Infodynamics for exploratory analysis of classification performance in Machine Learning. Infodynamics is an analogue of Thermodynamics for dealing with quantity of information instead of quantity of energy.

The premise is that if the information related to a random variable, the true class, wants to be "transported" somewhere to the predicted class, then the entropic balances of the true and predicted classes have to satisfy certain requisites.  


# Environment construction

```{r setup, include=FALSE}
library(dplyr)     # That infamous Mr. Wickham!
library(tidyr)     # Tidying tall & wide dataframes
library(infotheo)  # The functionality provided by this has to be rerouted through entropies
library(entropies) # This package
library(ggtern)    # Excellent package for ternary diagrams in the gg tradition
library(vcd)       # Categorical benchmarks
library(mlbench)   # ml benchmarkss
#knitr::opts_chunk$set(dev = 'pdf') # plots in pdf, better for publication
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment=NA, fig.width=6, fig.height=6)
fancy <- TRUE  # set this for nicer on-screen visualization
fancy <- FALSE # Set this for either printing matter or...

```

## Datasets available

```{r data-munging}
# the inventory of databases you can access
#library(datasets)
dsNames <- c("Ionosphere", "iris", "Glass", "Arthritis", "BreastCancer", "Sonar", "Wine") # 
className <- c("Class","Species", "Type", "Improved", "Class", "Class", "Cultivar")  # Name of class attribute
classVar <- c(35, 5, 10, 5, 11, 61, 1)   # ordinal of the class attribute
idNumber <- c(NaN, NaN, NaN, 1, 1, NaN, NaN) # Other attributes to dispose of: mainly identifiers.

K <- c(2, 3, 7, 3, 2, 2, 3)  # No. of classes in the class variable
# To select a dataset by name
# Caveat: you have to ensure that the containing package has been attached
evalDataset <- function(dsName){
    dsName <- as.character(dsName)
    switch(dsName, #TODO: improve this way of "invoking" the dataset.
        "iris" =         {data(iris); iris},
        "Ionosphere" =   {data(Ionosphere); Ionosphere},
        #Ionosphere$V2 is useless, V1 is highly correlated with Class
        "Glass" =        {data(Glass); Glass},
        "Arthritis" =    {data(Arthritis); Arthritis},
        "BreastCancer" = {data(BreastCancer); BreastCancer},
        "Sonar" =        {data(Sonar); Sonar},
        "Wine" =         {data(Wine); Wine}
    ) #This value "FALLS THROUGH"
}
m <- sapply(dsNames, function(n){nrow(evalDataset(n))}) # no. of instances in the dataset
n <- sapply(dsNames, function(n){ncol(evalDataset(n))}) - 1 - as.numeric(!is.nan(idNumber)) # no. of features in the dataset.
datasets <- data.frame(name=dsNames, 
                       className, 
                       idNumber, 
                       K=as.integer(K), 
                       n=as.integer(n), 
                       m, 
                       stringsAsFactors=FALSE)
# #To select the #of column of the classc
# whichClass <- function(ds, className){which(colnames(evalDatasset(ds))==className)}
# #whichNumVar <-  function(r){whichClass(evalDataset(r$name), r$className)}
# cardinalClass <- function(ds, className){
#     length(unique(evalDataset(ds)[,className]))
# }
# classVar <-  mapply(whichClass, datasets$name, datasets$className)
# K <- mapply(cardinalClass, datasets$name, classVar)
# library(dplyr)
# datasets <- data.frame(name,className, classVar, K)
datasets
```

Let's print this information to latex:

```{r}
library(xtable)
ds4latexing <- datasets %>% dplyr::select(-className, -idNumber)
row.names(ds4latexing) <- NULL
names(ds4latexing) <- c("Dataset Name", "class card.", "num. features", "num. instances")
thisLatex <- xtable(ds4latexing, 
                    caption="Some datasets considered in this study",
                    label="tab:datasets")
align(thisLatex) <- xalign(thisLatex)
thisLatex
```

# Data exploration

## Obtaining the entropies

Obtain the entropies and some other data for plotting from all datasets.

<!-- # as per:  -->
<!-- # http://stackoverflow.com/questions/5234117/how-to-drop-columns-by-name-in-a-data-frame -->
<!-- # Don't EVER use subset in PROGRAMS! -->
<!-- #ds <- subset(ds, subset=1:nrow(ds), select=dsRecord$className, drop=TRUE) -->

```{r find-entropies}
edf <- data.frame()
dsName <- "Ionosphere"#Testing and debugging purposes.
dsName <- "iris"
dsName <- "Glass"
#for(dsName in unique(datasets$name)){
    dsRecord <-  filter(datasets, name == dsName)
    ds <- evalDataset(dsName)
    print(sprintf("Analyzing dataset with class label: %s", dsName))
    Kdataset <- dplyr::select(ds, matches(dsRecord$className))
    Xdataset <- dplyr::select(ds, -matches(dsRecord$className))
    edf <- rbind(edf,
                cbind(dsName,
                    Transform="observation",
                    Vars="ALL",
                    jentropies(Kdataset,Xdataset)# entropies of observations
                )
    )
#                     sentropies(ds, nbins=ceiling(nrow(ds)^(1/3))) %>%
    #select those columns which are numeric
    numericColumns <- sapply(Xdataset, is.numeric)
    if (!any(numericColumns)) next #on categorical datasets no PCA, ICA, can be done
    Tdataset <- Xdataset[,numericColumns]#Until  we know how to apply Box-Cox
    # Tdataset <- log(Xdataset)#Box-Cox transformation
    # edf <- rbind(edf,
    #              cbind(dsName,
    #                    transform="Box-Cox:log",
    #                    jentropies(Xdataset,Tdataset)
    #                    )
    #              )
    pca <- prcomp(Tdataset, center = TRUE, scale. = TRUE)
    # print method
    print(pca)
    # plot method
    plot(pca, type = "l")
    PCAdataset <- as.data.frame(predict(pca, newdata=Tdataset)) #ICA data
    # Observe the PCA data by themselves
    for (i in 1:ncol(PCAdataset)){    
        edf <- rbind(edf,
                cbind(dsName,
                    Transform="PCA", #PCA_1 is the lable of
                    Vars=paste0(i),
                    jentropies(Xdataset,dplyr::select(PCAdataset, i))# only i-th PCA
                ),
                cbind(dsName,
                     Transform="PCAacc",
                     Vars=paste(1,i,sep="_"),#PCA_1 is the lable of
                    jentropies(Xdataset,dplyr::select(PCAdataset, (1:i)))# cum i-th PCA
                )
        )
    }
#}#for
str(edf)
# show the split entropies
#filter(edf, type != "XY")
#show all entropies for a data base
print(filter(edf, dsName == "iris"))
```

```{r visualization in CMET}
# show the split entropies
vetdf <- filter(edf, type != "XY") %>% 
    filter(grepl("PCA|PCAacc",Transform))
fancy <- TRUE
fancy <- FALSE
transfo_cmet <- ggmetern(vetdf, fancy) +
    geom_point(aes(color=Transform, shape=type), size=3) +
    scale_shape_manual(values=c(4, 20, 1)) +
    labs(color="transfo name", shape="Var type")# +
    #geom_label(mapping=aes(as.character(Vars)))#TODO: check with ggtern's website how TODO
transfo_cmet
```

Visualizing the PCA as a source of information:

```{r}
sedf <- 
```

