---
title: "Use case for the Multivariate Source Entropy Triangle"
author: "Francisco J. Valverde-Albacete & Carmen Peláez Moreno"
date: "14 de diciembre de 2015"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Environment construction

## Knitting options

```{r setup, include=FALSE}
#knitr::opts_chunk$set(dev = 'pdf') # plots in pdf, better for publication
knitr::opts_chunk$set(comment=NA, fig.width=6, fig.height=4)
knitr::opts_chunk$set(warning=FALSE)# Should not appear in the knitted document
```

## Library loading

```{r, echo=FALSE, warning=FALSE, environment}
library(tidyverse)  # Acceding to that infamous Mr. Wickham's requests!
library(infotheo)  # The functionality provided by this has to be rerouted through entropies
library(entropies) # This package
library(ggtern)    # Ternary diagrams on ggplot
library(vcd)       # Categorical benchmarks
library(mlbench)   # ml benchmarkss
library(candisc)   # Wine dataset
fancy <- TRUE  # set this for nicer on-screen visualization
#fancy <- FALSE # Set this for either printing matter
```

# An example of multivariate use

## Datasets available

The inventory of databases to be explored:

```{r dataset processing}
# the inventory of databases you can access
#library(datasets)
dsNames <- c("Ionosphere", "iris", "Glass", "Arthritis", "BreastCancer", "Sonar", "Wine") # 
className <- c("Class","Species", "Type", "Improved", "Class", "Class", "Cultivar")  # Name of class attribute
classVar <- c(35, 5, 10, 5, 11, 61, 1)   # ordinal of the class attribute
idNumber <- c(NaN, NaN, NaN, 1, 1, NaN, NaN) # Other attributes to dispose of: mainly identifiers.

K <- c(2, 3, 7, 3, 2, 2, 3)  # No. of classes in the class variable
# To select a dataset by name
# Caveat: you have to ensure that the containing package has been attached
evalDataset <- function(dsName){
    dsName <- as.character(dsName)
    switch(dsName, #TODO: improve this way of "invoking" the dataset.
        "iris" =         {data(iris); iris},
        "Ionosphere" =   {data(Ionosphere); Ionosphere},
        "Glass" =        {data(Glass); Glass},
        "Arthritis" =    {data(Arthritis); Arthritis},
        "BreastCancer" = {data(BreastCancer); BreastCancer},
        "Sonar" =        {data(Sonar); Sonar},
        "Wine" =         {data(Wine); Wine}
    ) #This value "FALLS THROUGH"
}
m <- sapply(dsNames, function(n){nrow(evalDataset(n))}) # no. of instances in the dataset
n <- sapply(dsNames, function(n){ncol(evalDataset(n))}) - 1 - as.numeric(!is.nan(idNumber)) # no. of features in the dataset.
datasets <- data.frame(name=dsNames, 
                       className, 
                       idNumber, 
                       K=as.integer(K), 
                       n=as.integer(n), 
                       m, 
                       stringsAsFactors=FALSE)
# #To select the #of column of the classc
# whichClass <- function(ds, className){which(colnames(evalDatasset(ds))==className)}
# #whichNumVar <-  function(r){whichClass(evalDataset(r$name), r$className)}
# cardinalClass <- function(ds, className){
#     length(unique(evalDataset(ds)[,className]))
# }
# classVar <-  mapply(whichClass, datasets$name, datasets$className)
# K <- mapply(cardinalClass, datasets$name, classVar)
# library(dplyr)
# datasets <- data.frame(name,className, classVar, K)
datasets
```

Let's print this information to latex:

```{r datasets to latex}
library(xtable)
ds4latexing <- datasets %>% select(-className, -idNumber)
row.names(ds4latexing) <- NULL
names(ds4latexing) <- c("Dataset Name", "class card.", "num. features", "num. instances")
thisLatex <- xtable(ds4latexing, 
                    caption="Some datasets considered in this study",
                    label="tab:datasets")
align(thisLatex) <- xalign(thisLatex)
thisLatex
```

## Obtaining the entropies

Obtain the entropies and some other data for plotting from all datasets.

```{r find-entropies, warning=FALSE}
edf <- data.frame()
for(i in 1:nrow(datasets)){
    dsRecord <-  datasets[i, ] #filter(datasets, name == dsName)
    dsName <- dsRecord$name
    edf <- rbind(
        edf,
        getDatasetSourceEntropies(evalDataset(dsName), 
            dsName, 
            className = dsRecord$className,
            withClass=TRUE, 
            type="total")
        )
}
str(edf)
```


## Plotting the aggregate data

Now we print the total entropy balance in the triangle for the different datasets. This allows us to compare the different relative compositions of the datasets. 

```{r, aggregate-SMET}
#excised <- c("name", "withClass", "isClass", "dsName") #Or any other that we add
totalEdf <-  edf %>% 
            mutate(name=dsName) %>%
            group_by(name, withClass) %>%
            summarise(H_Ux = sum(H_Uxi),
                    H_Px = sum(H_Pxi),
                    DeltaH_Px = sum(DeltaH_Pxi),
                    VI_Px = sum(VI_Pxi),
                    M_Px = sum(M_Pxi))
           
#totalEdf
# Plot the triangle with only the aggregate data with no Class
smet <-  ggmetern(totalEdf %>% filter(withClass == FALSE), fancy) + 
    geom_point(mapping=aes(shape=name), size=3)+
    scale_shape_manual(values=1:nrow(datasets)) +
    labs(shape="Dataset")
if (fancy){
    smet <- smet + ggtitle("Source Multivariate Entropy by dataset")
}
smet
ggsave("aggregated_without_label.jpeg", plot=smet)

# A color blind palette from: http://www.cookbook-r.com/Graphs/Colors_%28ggplot2%29/#a-colorblind-friendly-palette
# The palette with grey:
#cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# The palette with black:
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

#Use the dsName for shape and the withClass for fill => use only fillable glyphs.
smet <-  ggmetern(totalEdf, fancy) + 
    geom_point(mapping=aes(shape=name, color=withClass), size=3) +
    scale_colour_manual(values=cbbPalette) +
    #scale_color_manual(values=c("red", "black")) +
    scale_shape_manual(values=1:nrow(datasets)) +
    labs(shape="Dataset", colour="Using class")
if (fancy){
    smet <- smet + ggtitle("Source Multivariate Entropy by dataset")
}
smet
ggsave("aggregated_withWo_label.jpeg", plot=smet)
```

## Observing the entropy balances in the class features

```{r}
onlyClasses <- edf %>% filter(isClass==TRUE)
# Plot the triangle with only the aggregate data with no Class
smetClasses <-  ggmetern(onlyClasses, fancy) + 
    geom_point(mapping=aes(shape=dsName), size=3) +
    scale_shape_manual(values=1:nrow(onlyClasses)) +
    labs(shape="Dataset")
if (fancy){
    smetClasses <- smetClasses + ggtitle("Source Multivariate Entropy for Classes accross datasets")
}
smetClasses
ggsave("multisplit_for_class_features.jpeg", plot=smetClasses)
```

## Plotting the multisplit data

We choose some of the interesting datasets from the diagram above to investigate:

```{r, specific-data}
#TODO: make a grid of these plots to be able to see anything different. 
thisDsName <- "Ionosphere" # CAVEAT! Not enough different glyphs!!!
thisDsName <- "iris" # for paper, first run this value, then with "Glass"
# thisDsName <- "Glass"
# thisDsName <- "Arthritis"
# thisDsName <- "BreastCancer"
# thisDsName <- "Sonar"
# thisDsName <- "Wine"
# negatively subsetting recipe from Stack Overflow
thisEdf <-  rbind(edf %>% filter(dsName == thisDsName & name != "ALL") %>% 
                      select(-starts_with("isClass")), #%>% 
                      #filter(name != "ALL"),
                  ungroup(totalEdf) %>% filter(name == thisDsName) %>% 
                      mutate(dsName=name, name = "@AGGREGATE") %>%
                      select(name, 
                             H_Uxi=H_Ux, H_Pxi=H_Px, DeltaH_Pxi=DeltaH_Px,
                             M_Pxi=M_Px, 
                             VI_Pxi=VI_Px,
                             withClass, dsName)
                ) 
# Create different geometris for different feature set cardinalities:
# First consider the features in the dataset without the class variable
thisEt <-  ggmetern(filter(thisEdf, withClass == FALSE),  fancy) 
if ((nrow(thisEdf) - 1)/2 > 14){#too many points to be represented with glpsh
    thisEt <- thisEt + #geom_density_tern(aes(fill=..level..)) +
        stat_density_tern(geom='polygon',
                        aes(fill=..level..),
                        #base=base,  ###NB Base Specification
                        colour='grey50') + 
        scale_fill_gradient(low='green',high='red')  +
        geom_point(size=1)
}else {
    thisEt <- thisEt + geom_point(aes(shape=name), size=3) +
        scale_shape_manual(values=1:14) + 
        labs(shape="Feature") #+
    #ggtitle("Source Multivariate Entropies per Feature")
}
thisEt
ggsave(filename=sprintf("%s_without_class.jpeg", thisDsName))

# Now consider the same set with the class label included.
thisEt <-  ggmetern(thisEdf, fancy) 
if ((nrow(thisEdf) - 1)/2 > 14){#too many points to be represented with glpsh
    thisEt <- thisEt + #geom_density_tern(aes(fill=..level..)) +
        stat_density_tern(geom='polygon',
                        aes(fill=..level..),
                        #base=base,  ###NB Base Specification
                        colour='grey50') + 
        scale_fill_gradient(low='green',high='red')  +
        geom_point(size=1)
}else {
    thisEt <- thisEt + geom_point(aes(shape=name, colour=withClass), size=3) +
        scale_shape_manual(values=1:14) + 
        #scale_color_discrete("color_blind")
        scale_colour_manual(values=cbbPalette)
    #ggtitle("Source Multivariate Entropies per Feature")
}
thisEt + labs(shape="Feature", color="Using class") #+
ggsave(filename=sprintf("%sW_WO_class.jpeg", thisDsName))
```


### The absolute decomposition

```{r stacked-bars}
# In case the switch is ON for excluding the aggregate.
excludeAggregate <- TRUE
#excludeAggregate <- FALSE

analyzeWithClass <- FALSE
#analyzeWithClass <- TRUE
if (analyzeWithClass){
    smedf <- filter(thisEdf, withClass) %>% select(-withClass)
} else {
# For this once let's just use the entropy with no class variable
    smedf <- filter(thisEdf, !withClass) %>% select(-withClass) # source multivariate entropy data frame
}
#library(scales)
p <-  ggmebars(smedf, excludeAggregate,proportional=FALSE)
p + ylab("Source Multivariate Entropy") + xlab("Feature/Variable") #+
    #ggtitle("Absolute Source Multivariate Entropies per Feature")
    # scale_y_continuous(trans=log2_trans(),
    #                 breaks=trans_breaks("log2", function(x) 2^x),
    #                 labels=trans_format("log2", math_format(2^.x)))
# 
ggsave(filename=sprintf("%s_entropy_bars_noAgg.jpeg", thisDsName))
```


The multisplit SMET can also offer the same information with a different encoding: using the fill color on the glyphs used to represent the maximum entropy per feature:

```{r absolute-SMET}
ggmetern(filter(smedf, name != "@AGGREGATE"), fancy) +
    geom_point(size=4, aes(shape=name, colour=H_Uxi)) + 
    scale_shape_manual(values=1:14, guide=FALSE) + 
    scale_color_gradient(low="grey", high="black") +
    labs(shape="Feature", colour="$\\textit{H_{U_{X_i}}}") #+
    #ggtitle("Absolute Source Multivariate Entropies per Feature") #+     theme(legend.position="bottom")
    #geom_point(size=4, aes(shape=name, colour=withClass))
ggsave(filename=sprintf("%s_smet_noAgg_absoluteEntropy.jpeg", thisDsName))
```


However, turning the stacked bar graph into a pie chart is a bad idea, since the remaining information $VI_{P_{X_i}}$ is de-emphasized (that is, compare with the area in the stacked bar graph above). To a certain extent, this is equivalent to the Source Multivariate Entropy Triangle:

```{r entropy-pie}
 p + ylab("Source Multivariate Entropy") + xlab("Feature/Variable") + coord_polar() #+
    #ggtitle("Relative Source Multivariate Entropies per Feature")
ggplot2::ggsave(filename=sprintf("%s_entropy_pie_noAgg.jpeg", thisDsName))
```

# Postscriptum

```{r ps}
sessionInfo()
```
