---
title: "Use case for the Multivariate Source Entropy Triangle"
author: "Francisco J. Valverde-Albacete & Carmen Peláez Moreno"
date: "14 de diciembre de 2015"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Environment construction

```{r, echo=FALSE, environment}
#library(ggtern)   # Excellent package for ternary diagrams in the gg tradition
#library(entropy)  # To work out the appropriate coordinates.
library(dplyr)     # That infamous Mr. Wickham!
library(tidyr)     # Tidying tall & wide dataframes
library(infotheo)  # The functionality provided by this has to be rerouted through entropies
library(entropies) # This package
library(ggtern)    # Ternary diagrams on ggplot
library(vcd)       # Categorical benchmarks
library(mlbench)   # ml benchmarkss
library(candisc)   # Wine dataset
#knitr::opts_chunk$set(dev = 'pdf') # plots in pdf, better for publication
knitr::opts_chunk$set(comment=NA, fig.width=6, fig.height=4)
fancy <- TRUE  # set this for nicer on-screen visualization
fancy <- FALSE # Set this for either printing matter
```


# An example of multivariate use

## Datasets available

The inventory of databases to be explored:

```{r data-munging}
# the inventory of databases you can access
#library(datasets)
dsNames <- c("Ionosphere", "iris", "Glass", "Arthritis", "BreastCancer", "Sonar", "Wine") # 
className <- c("Class","Species", "Type", "Improved", "Class", "Class", "Cultivar")  # Name of class attribute
classVar <- c(35, 5, 10, 5, 11, 61, 1)   # ordinal of the class attribute
idNumber <- c(NaN, NaN, NaN, 1, 1, NaN, NaN) # Other attributes to dispose of: mainly identifiers.

K <- c(2, 3, 7, 3, 2, 2, 3)  # No. of classes in the class variable
# To select a dataset by name
# Caveat: you have to ensure that the containing package has been attached
evalDataset <- function(dsName){
    dsName <- as.character(dsName)
    switch(dsName, #TODO: improve this way of "invoking" the dataset.
        "iris" =         {data(iris); iris},
        "Ionosphere" =   {data(Ionosphere); Ionosphere},
        "Glass" =        {data(Glass); Glass},
        "Arthritis" =    {data(Arthritis); Arthritis},
        "BreastCancer" = {data(BreastCancer); BreastCancer},
        "Sonar" =        {data(Sonar); Sonar},
        "Wine" =         {data(Wine); Wine}
        )#This value "FALLS THROUGH"
}
m <- sapply(dsNames, function(n){nrow(evalDataset(n))}) # no. of instances in the dataset
n <- sapply(dsNames, function(n){ncol(evalDataset(n))}) - 1 - as.numeric(!is.nan(idNumber)) # no. of features in the dataset.
datasets <- data.frame(name=dsNames, 
                       className, 
                       idNumber, 
                       K=as.integer(K), 
                       n=as.integer(n), 
                       m, 
                       stringsAsFactors=FALSE)
 
# #To select the #of column of the classc
# whichClass <- function(ds, className){which(colnames(evalDatasset(ds))==className)}
# #whichNumVar <-  function(r){whichClass(evalDataset(r$name), r$className)}
# cardinalClass <- function(ds, className){
#     length(unique(evalDataset(ds)[,className]))
# }
# classVar <-  mapply(whichClass, datasets$name, datasets$className)
# K <- mapply(cardinalClass, datasets$name, classVar)
# library(dplyr)
# datasets <- data.frame(name,className, classVar, K)
datasets
```

Let's print this information to latex:

```{r}
library(xtable)
ds4latexing <- datasets %>% select(-className, -idNumber)
row.names(ds4latexing) <- NULL
names(ds4latexing) <- c("Dataset Name", "class card.", "num. features", "num. instances")
thisLatex <- xtable(ds4latexing, 
                    caption="Some datasets considered in this study",
                    label="tab:datasets")
align(thisLatex) <- xalign(thisLatex)
thisLatex
```

## Obtaining the entropies

Obtain the entropies and some other data for plotting from all datasets.

```{r find-entropies}
# Analyze both with Classes and without
withClasses <- c(TRUE, FALSE)
#  "name"       "H_Uxi"      "H_Pxi"      "DeltaH_Pxi" "VI_Pxi"     "M_Pxi" 
# "withClass", "dsName", "isClass"
edf <- data.frame()
for(i in 1:nrow(datasets)){
    dsRecord <-  datasets[i, ] #filter(datasets, name == dsName)
    dsName <- dsRecord$name
    ds <- infotheo::discretize( # Controlled discretization of the database.
        evalDataset(dsName), 
        nbins=max(ceiling(nrow(Glass)^(1/2)), dsRecord$K)# At least the classes.
    )
    if (!is.nan(dsRecord$idNumber)) # ids as features: spureously correlated with class labels
        ds <- ds[,-dsRecord$idNumber]
    for(withClass in withClasses){
        if (withClass){
            print(sprintf("Analyzing dataset with class label: %s", dsName))
        }else {
            print(sprintf("Analyzing dataset without class label: %s", dsName))
            # as per: 
            # http://stackoverflow.com/questions/5234117/how-to-drop-columns-by-name-in-a-data-frame
            # Don't EVER use subset in PROGRAMS!
            #ds <- subset(ds, subset=1:nrow(ds), select=dsRecord$className, drop=TRUE)
            ds <- ds[, !colnames(ds) == dsRecord$className] #fastest in bechnmark at bot. of url
        }
        edf <- rbind(edf,
                     sentropies(ds) %>% 
                         mutate(withClass = withClass,
                                isClass = (as.character(dsRecord$className) ==
                                                as.character(name)),
                                dsName = dsName
                         )
        )
    }
}
str(edf)
```


## Plotting the aggregate data

Now we print the total entropy balance in the triangle for the different datasets. This allows us to compare the different relative compositions of the datasets. 

```{r, aggregate-SMET}
#excised <- c("name", "withClass", "isClass", "dsName") #Or any other that we add
totalEdf <-  edf %>% 
            mutate(name=dsName) %>%
            group_by(name, withClass) %>%
            summarise(H_Uxi = sum(H_Uxi),
                    H_Pxi = sum(H_Pxi),
                    DeltaH_Pxi = sum(DeltaH_Pxi),
                    VI_Pxi = sum(VI_Pxi),
                    M_Pxi = sum(M_Pxi))
           
totalEdf
# Plot the triangle with only the aggregate data with no Class
smet <-  ggmetern(totalEdf %>% filter(withClass == FALSE), fancy) + 
    geom_point(mapping=aes(shape=name), size=3)+
    scale_shape_manual(values=1:nrow(datasets)) +
    labs(shape="Dataset")
if (fancy){
    smet <- smet + ggtitle("Source Multivariate Entropy by dataset")
}
smet
ggsave("aggregated_without_label.jpeg", plot=smet)

#Use the dsName for shape and the withClass for fill => use only fillable glyphs.
smet <-  ggmetern(totalEdf, fancy) + 
    geom_point(mapping=aes(shape=name, color=withClass), size=3)+
    scale_color_manual(values=c("red", "black")) +
    scale_shape_manual(values=1:nrow(datasets)) +
    labs(shape="Dataset", colour="Using class")
if (fancy){
    smet <- smet + ggtitle("Source Multivariate Entropy by dataset")
}
smet
ggsave("aggregated_withWo_label.jpeg", plot=smet)
```

## Plotting the multisplit data

We choose some of the interesting datasets from the diagram above to investigate:

```{r, specific-data}
#TODO: make a grid of these plots to be able to see anything different. 
thisDsName <- "Ionosphere" # CAVEAT! Not enough different glyphs!!!
thisDsName <- "iris" # for paper, first run this value, then with "Glass"
# thisDsName <- "Glass"
# thisDsName <- "Arthritis"
# thisDsName <- "BreastCancer"
# thisDsName <- "Sonar"
# thisDsName <- "Wine"
# negatively subsetting recipe from Stack Overflow
thisEdf <-  rbind(edf %>% filter(dsName == thisDsName & name != "ALL") %>% 
                      select(-starts_with("isClass")), #%>% 
                      #filter(name != "ALL"),
                  ungroup(totalEdf) %>% filter(name == thisDsName) %>% 
                      mutate(dsName=name, name = "@AGGREGATE") %>%
                      select(name, 
                             H_Uxi, H_Pxi, DeltaH_Pxi, M_Pxi, VI_Pxi,
                             withClass, dsName)
                ) 
# Create different geometris for different feature set cardinalities:
# First consider the features in the dataset without the class variable
thisEt <-  ggmetern(filter(thisEdf, withClass == FALSE),  fancy) 
if ((nrow(thisEdf) - 1)/2 > 14){#too many points to be represented with glpsh
    thisEt <- thisEt + #geom_density_tern(aes(fill=..level..)) +
        stat_density_tern(geom='polygon',
                        aes(fill=..level..),
                        #base=base,  ###NB Base Specification
                        colour='grey50') + 
        scale_fill_gradient(low='green',high='red')  +
        geom_point(size=1)
}else {
    thisEt <- thisEt + geom_point(aes(shape=name), size=3) +
        scale_shape_manual(values=1:14) + 
        labs(shape="Feature") #+
    #ggtitle("Source Multivariate Entropies per Feature")
}
thisEt
ggsave(filename=sprintf("%s_without_class.jpeg", thisDsName))

# Now consider the same set with the class label included.
# A color blind palette from: http://www.cookbook-r.com/Graphs/Colors_%28ggplot2%29/#a-colorblind-friendly-palette
# The palette with grey:
#cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# The palette with black:
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
thisEt <-  ggmetern(thisEdf, fancy) 
if ((nrow(thisEdf) - 1)/2 > 14){#too many points to be represented with glpsh
    thisEt <- thisEt + #geom_density_tern(aes(fill=..level..)) +
        stat_density_tern(geom='polygon',
                        aes(fill=..level..),
                        #base=base,  ###NB Base Specification
                        colour='grey50') + 
        scale_fill_gradient(low='green',high='red')  +
        geom_point(size=1)
}else {
    thisEt <- thisEt + geom_point(aes(shape=name, colour=withClass), size=3) +
        scale_shape_manual(values=1:14) + 
        #scale_color_discrete("color_blind")
        scale_colour_manual(values=cbbPalette)
    #ggtitle("Source Multivariate Entropies per Feature")
}
thisEt + labs(shape="Feature", color="Using class") #+
ggsave(filename=sprintf("%sW_WO_class.jpeg", thisDsName))
```


### The absolute decomposition

```{r stacked-bars}
# In case the switch is ON for excluding the aggregate.
excludeAggregate <- TRUE
#excludeAggregate <- FALSE

analyzeWithClass <- FALSE
#analyzeWithClass <- TRUE
if (analyzeWithClass){
    smedf <- filter(thisEdf, withClass) %>% select(-withClass)
} else {
# For this once let's just use the entropy with no class variable
    smedf <- filter(thisEdf, !withClass) %>% select(-withClass) # source multivariate entropy data frame
}
#library(scales)
p <-  ggmebars(smedf, excludeAggregate,proportional=FALSE)
p + ylab("Source Multivariate Entropy") + xlab("Feature/Variable") #+
    #ggtitle("Absolute Source Multivariate Entropies per Feature")
    # scale_y_continuous(trans=log2_trans(),
    #                 breaks=trans_breaks("log2", function(x) 2^x),
    #                 labels=trans_format("log2", math_format(2^.x)))
# 
```


The multisplit SMET can also offer the same information with a different encoding: using the fill color on the glyphs used to represent the maximum entropy per feature:

```{r absolute-SMET}
ggmetern(filter(smedf, name != "@AGGREGATE"), fancy) +
    geom_point(size=4, aes(shape=name, colour=H_Uxi)) + 
    scale_shape_manual(values=1:14, guide=FALSE) + 
    scale_color_gradient(low="grey", high="black") +
    labs(shape="Feature", colour="$\\textit{H_{U_{X_i}}}") #+
    #ggtitle("Absolute Source Multivariate Entropies per Feature") #+     theme(legend.position="bottom")
    #geom_point(size=4, aes(shape=name, colour=withClass))
ggsave(filename=sprintf("%s_smet_noAgg_absoluteEntropy.jpeg", thisDsName))
```


However, turning the stacked bar graph into a pie chart is a bad idea, since the remaining information $VI_{P_{X_i}}$ is de-emphasized (that is, compare with the area in the stacked bar graph above). To a certain extent, this is equivalent to the Source Multivariate Entropy Triangle:

```{r entropy-pie}
 p + ylab("Source Multivariate Entropy") + xlab("Feature/Variable") + coord_polar() #+
    #ggtitle("Relative Source Multivariate Entropies per Feature")
ggplot2::ggsave(filename=sprintf("%s_entropy_pie_noAgg.jpeg", thisDsName))
```

# Postscriptum

```{r ps}
sessionInfo()
```

<!-- # SO FAR SEEMS OK -->
<!-- ## Facetting on different values -->


<!-- ```{r} -->
<!-- selected <- c("iris", "Glass", "BreastCancer") -->
<!-- #selEdf <- edf %>% filter(dsName %in% selected) -->
<!-- thisSelEdf <-  rbind(edf %>% filter(dsName %in% selected) %>%select(-starts_with("isClass")), -->
<!--                   totalEdf %>% filter(dsName %in% selected) %>% mutate(name = "AGGREGATE") %>% -->
<!--                       select(name, #matches(".*"),  -->
<!--                              H_Uxi = H_Ux, H_Pxi = H_Px, DeltaH_Pxi = DeltaH_Px, -->
<!--                              VI_Pxi = VI_Px, M_Pxi = M_Px, -->
<!--                              withClass, dsName) -->
<!--                 ) -->
<!-- #edf1 <- entropies(df, nbins=ceiling(nrow(df)^(1/3))) -->
<!-- #edf1 -->
<!-- thisSelEt <-  ggmetern(thisEdf, size=4, aes(shape=name)) #, colour=withClass)) -->
<!-- thisSelEt + facet_grid(dsName ~ withClass) +  -->
<!--     scale_colour_manual(values = c("black", "red")) + -->
<!--     scale_shape_manual(values=1:14) +  -->
<!--     labs(shape="feature") #s, colour="Using class") -->
<!-- ``` -->

<!-- # Use on different datasets -->

<!-- Get some categorical datasets from package "vcd" -->
<!-- ```{r} -->
<!-- library(vcd) -->
<!-- library(mlbench) -->
<!-- pname <-  "vcd" # categorical variables -->
<!-- pname <- "mlbench" -->
<!-- ps  <- data(package=pname)     # get all the info for the data in the package -->
<!-- vcdData <- ps$results[,"Item"] # get the name of all the data  -->
<!-- #Now find the names of the datasets that are categorical! -->
<!-- vcdDataframes  <- vcdData[unlist(lapply(ps$results[,"Item"], -->
<!--                                         function(x){is.data.frame(eval(as.name(x)))}) -->
<!--                                  ) -->
<!--                           ] -->
<!-- #list them -->
<!-- vcdDataframes -->
<!-- ``` -->

<!-- Do the same treatment on Arthritis: -->

<!-- ```{r} -->
<!-- withClass <-  TRUE -->
<!-- withClass <-  FALSE -->
<!-- if (withClass){# Consider the class attribute -->
<!--     df <- iris -->
<!--     #df <- Arthritis -->
<!--     #df <- BreastCancer[,-1] # Take away identifiers. -->
<!--     df <- {data(Glass); Glass} -->
<!--     #df <- {data(DNA); DNA} -->
<!-- }else{ #Do not consider the class attribute or e.g. identities. -->
<!--     df <- iris[, -c(5)] -->
<!--     #df <- Arthritis[,c(2:4)] -->
<!--     #df <- BreastCancer[,-c(1,11)] # 1 is an identifier, 11 is the Class -->
<!--     df <- Glass[,-c(10)] # 10 = Type and identifier of the type of Glass! -->
<!--     #df <- DNA[,-c(181)] # V. 181 = Class -->
<!-- } -->
<!-- edf <- entropies(df, nbins=ceiling(nrow(df)^(1/3))) -->
<!-- # Number one is always the name of the features! -->
<!-- addTotalEntropies <- function(edf, label="@Total"){ -->
<!--   edf <-cbind( -->
<!--     data.frame(name=c(as.character(edf[,1]), label)), -->
<!--     rbind(edf[,-1], colSums(edf[,-1])) -->
<!--     ) -->
<!--   return(edf) -->
<!-- } -->
<!-- edf <- addTotalEntropies(edf) -->
<!-- ep <-  ggmetern(edf) + labs(shape="Feature") -->
<!-- ep -->
<!-- ``` -->
