---
title: "Infodynamics of classification"
author: "Francisco J. Valverde-Albacete & Carmen Peláez Moreno"
date: "9 de mayo de 2017"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# Introduction

This vignette tries to demonstrate the use of Infodynamics for exploratory analysis of classification performance in Machine Learning. Infodynamics is an analogue of Thermodynamics for dealing with quantity of information instead of quantity of energy.

The premise is that if the information related to a random variable, the true class, wants to be "transported" somewhere to the predicted class, then the entropic balances of the true and predicted classes have to satisfy certain requisites.  


# Environment construction

## Knitting options

```{r setup, include=FALSE}
#knitr::opts_chunk$set(dev = 'pdf') # plots in pdf, better for publication
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_chunk$set(dev = 'pdf') # plots in pdf, better for publication
knitr::opts_chunk$set(comment=NA, fig.width=6, fig.height=4)
knitr::opts_chunk$set(warning=FALSE)# Should not appear in the knitted document
```

## Library loading

```{r libraries, include=FALSE}
library(tidyverse)  # Acceding to that infamous Mr. Wickham's requests!
# library(dplyr)     # That infamous Mr. Wickham!
# library(tidyr)     # Tidying tall & wide dataframes
library(infotheo)  # The functionality provided by this has to be rerouted through entropies
library(ggtern)    # Excellent package for ternary diagrams in the gg tradition
library(entropies) # This package
library(caret)     # General framework for classification and regression
library(e1071)     # Many facilities and classifiers
library(vcd)       # Categorical benchmarks
library(mlbench)   # ml benchmarkss
library(randomForest) # random forest classifiers
library(candisc)   # Wine dataset
```
## Global parameter definitions

```{r}
fancy <- TRUE  # set this for nicer on-screen visualization
fancy <- FALSE 
inDebugMode <- TRUE # general switch for debugging
#inDebugMode <- FALSE
writePlots <- TRUE # A switch to write the plots
writePlots <- FALSE 
```

# Data preparation

## Datasets available

```{r explicit-datasets}
datasets <- getDatasets()
library(xtable)
ds4latexing <- datasets %>% dplyr::select(-className, -idNumber)
row.names(ds4latexing) <- NULL
names(ds4latexing) <- c("Dataset Name", "class card.", "num. features", "num. instances")
thisLatex <- xtable(ds4latexing, 
                    caption="Some datasets considered in this study",
                    label="tab:datasets")
align(thisLatex) <- xalign(thisLatex)
thisLatex
```

# Data exploration

## Obtaining the entropies

Obtain the entropies and some other data for plotting from all datasets.

<!-- # as per:  -->
<!-- # http://stackoverflow.com/questions/5234117/how-to-drop-columns-by-name-in-a-data-frame -->
<!-- # Don't EVER use subset in PROGRAMS! -->
<!-- #ds <- subset(ds, subset=1:nrow(ds), select=dsRecord$className, drop=TRUE) -->

```{r find-entropies}
#method = 'rf'
# Naive transformation from factors to numbers in 0 to num.factors - 1
factor.as.numeric <- function(f){
    nums <- as.numeric(f)
    return(nums - min(nums))
}
cmet_edf <- tibble()# Gathers the CMET entropies
smet_edf <- tibble()# Gathers the SMET entropies
dsNames <- unique(datasets$name)#debug
dsNames <-  dsNames[1]#debug
#dsNames <- dsNames[c(2,4)]#Select iris, Arthritis some of them for easier observation.
#dsNames <- dsNames[c(3,5)]#Select Glass, BreastCancer some of them for easier observation.
dsName <- dsNames[1]#Select Ionosphere
#dsName <- dsNames[2] # iris
#dsNames <- dsNames[3]
#dsNames <- dsNames[4]
#dsNames <- dsNames[5]
for(dsName in dsNames){
    dsRecord <-  filter(datasets, name == dsName)
    ds <- evalDataset(dsName) %>% as_tibble()
    if (is.numeric(dsRecord$idNumber) & !(is.nan(dsRecord$idNumber))){
        ds <- dplyr::select_(ds, -dsRecord$idNumber)
    }
    nbins <- ceiling(dsRecord$m^(1/3))# This is the default!
    #ncols <- min(10, ncol(ds))
#    for(withClass in withClasses){
#        if (withClass){
            print(sprintf("Analyzing dataset with class label: %s", dsName))
#        }else {
#            print(sprintf("Analyzing dataset without class label: %s", dsName))
            # as per: 
            # http://stackoverflow.com/questions/5234117/how-to-drop-columns-by-name-in-a-data-frame
            # Don't EVER use subset in PROGRAMS!
            #ds <- subset(ds, subset=1:nrow(ds), select=dsRecord$className, drop=TRUE)
#            ds <- ds[, !colnames(ds) == dsRecord$className] #fastest in bechnmark at bot. of url
#        }
        #Kdataset <- select(ds, which(names(ds) == dsRecord$className))
        # 0. Select class labels and carry out stratified partitioning
        Kdataset <- dplyr::select(ds, matches(dsRecord$className))
        names(Kdataset) <- "Class" #This is very brittle
        inTrain <- createDataPartition(
            y = Kdataset$Class, ##outcome data are needed
            p = .75,      ##The percentage of data in the training set
            list = FALSE  ## The format of the results
            )
        smet_edf <- rbind(smet_edf, 
                          cbind( dsName, group="labels", phase="training",
                                 sentropies(Kdataset[inTrain,],type="dual")),
                          cbind( dsName, group="labels", phase="testing",
                                 sentropies(Kdataset[-inTrain,],type="dual")))
        # 1. Select the observations and do some cleaning
        Xdataset <- dplyr::select(ds, -matches(dsRecord$className))
        Xdataset <- Xdataset %>%     
            #transform factors to number
            mutate_if(is.factor,factor.as.numeric) %>%
            # Dispose of constant columns: they carry no information
            select_if(function(v)(var(v) > 0)) %>% 
            # Dispose of columns with NaN
            select_if(function(v) !any(is.na(v)))
        # Try to discretize before calculating entropies so as to avoid
        # iterated discretizations in primitives. 
        # Discretize to have a standard set to carry out entropies.
        dXdataset <- infotheo::discretize(Xdataset, disc="equalwidth",nbins)
        # CAVEAT: we also need it in numeric form for transformations!
        smet_edf <- rbind(smet_edf, 
                        cbind(dsName,group="observations", phase="training",
                              sentropies(dXdataset[inTrain,],type="dual")),
                        cbind(dsName,group="observations", phase="testing",
                              sentropies(dXdataset[-inTrain,],type="dual")))
        cmet_edf <- rbind(cmet_edf,
                     cbind(dsName,
                           transform="observation", phase="training",
                           jentropies(Kdataset[inTrain,],dXdataset[inTrain,])),
                     cbind(dsName,
                           transform="observation", phase="testing",
                           jentropies(Kdataset[-inTrain,],dXdataset[-inTrain,]))
                     )
        # numericColumns <- sapply(Xdataset, is.numeric)
        # if (!any(numericColumns)) next#improve this for classifications
        # #catColumns <- !numericColumns#categorical columns in X (not including the class)
        # Tdataset <- Xdataset[,numericColumns]#Until  we know how to apply Box-Cox
        #select those columns which are numeric
        #Postcondition:
        #all(sapply(Tdataset, is.numeric))
        preTransform <- ""#no transformation
        # Tdataset <- log(Xdataset)#Box-Cox transformation
        # cmet_edf <- rbind(cmet_edf,
        #              cbind(dsName,
        #                    transform="Box-Cox:log",
        #                    jentropies(Xdataset,Tdataset)
        #                    )
        #              )
        pcaModel <- prcomp(Xdataset, center = TRUE, scale. = TRUE)
        plot(pcaModel, type = "l")
        Fvector <- as.tibble(predict(pcaModel, newdata=Xdataset))
        dFvector <- infotheo::discretize(Fvector,disc="equalwidth")
        #TODO: we should be printing how the variance accrues, etc
        postTransform <- "PCA"
        # Ydataset <- cbind(
        #                 Xdataset[,!numericColumns],
        #                 predict(pca, newdata=Tdataset)
        # )
        #### Do some sort of analysis on FVector and FVectorDiscretized
        for(i in (1:(min(10, ncol(ds))))){
            #first accrue the 
            cmet_edf <- rbind(cmet_edf,
                          cbind(dsName=paste0("1_",i), 
                                transform="PCA",
                                phase="training",
                                jentropies(dXdataset[inTrain,],
                                           as.tibble(dFvector[inTrain,1:i]))),
                          cbind(dsName=paste0("1_",i), 
                                transform="PCA",
                                phase="testing",
                                jentropies(dXdataset[-inTrain,],
                                           as.tibble(dFvector[-inTrain,1:i])))
                          )
            smet_edf <- rbind(smet_edf,
                    cbind(dsName = paste0("1_", i),
                          group="features", phase="training",
                          sentropies(as.tibble(dFvector[inTrain,1:i]),
                                     type="dual")),
                    cbind(dsName = paste0("1_", i),
                          group="features", phase="testing",
                          sentropies(as.tibble(dFvector[-inTrain,1:i]),
                                     type="dual"))
                    )
        }#Go over all possible nums of cols.
        # cmet_edf <- rbind(cmet_edf,
        #              cbind(dsName, 
        #                    transform=paste0(postTransform,":",preTransform),
        #                    phase="training",
        #                    jentropies(dXdataset[inTrain,], Fvector[inTrain,])
        #                    ),
        #              cbind(dsName, 
        #                    transform=paste0(postTransform,":",preTransform),
        #                    phase="testing",
        #                    jentropies(dXdataset[-inTrain,], Fvector[-inTrain,])
        #                    )
        # )
        # smet_edf <- rbind(smet_edf, 
        #                 cbind(dsName,group="features", phase="training",
        #                       sentropies(Fvector[inTrain,],type="dual")),
        #                 cbind(dsName,group="features", phase="testing",
        #                       sentropies(Fvector[-inTrain,],type="dual"))
        #                 )
        # inTrain <- createDataPartition(
        #     y = newDataset$Class, ##outcome data are needed
        #     p = .75,      ##The percentage of data in the training set
        #     list = FALSE  ## The format of the results
        #     )
                    # Finally, classification with some technique
        set.seed(107)
        #dFvector <- as.tibble(infotheo::discretize(Fvector, disc="equalwidth"))
        newDataset <- cbind(Kdataset[1],dFvector)
        names(newDataset)[1] <- "Class"
        ctrl <- trainControl(method = "repeatedcv",
                             repeats = 3)
        trainCM <- list() # Will keep the Conf Matrices of several classifiers
        testCM <- list()
        for( i in (1:(min(10, ncol(ds))))){
            training <- newDataset[ inTrain, 1:(i+1)]
            testing <- newDataset[ -inTrain, 1:(i+1)]
            # if (inDebugMode){ # Seeing what the partition is
            #     nrow(training)
            #     nrow(testing)
            # }
            #estimating the model
            print(sprintf("Estimating model with %d features for %s", 
                          i, dsName))
            rfModel <- train(Class ~ ., 
                data = training,
                method = "rf",
                trControl = ctrl)
            hatKdataset <- predict(rfModel, newdata = newDataset[ ,1:(i+1)]) %>% 
                as.tibble()
            #hatKdataset <- predict(rfModel, newdata = testing) %>% as.tibble()
            #if (inDebugMode){#These are the end to end measures, but inverted
            # data: is the PREDICTED data.
            # the data are in trainCM[[i]]$table
            trainCM[[i]] <- confusionMatrix(data = hatKdataset[inTrain,]$value, 
                                       training$Class)
            testCM[[i]] <- confusionMatrix(data = hatKdataset[-inTrain,]$value,
                                      testing$Class)
            #}
            #Now do the ID analysis on the predicted data.
            cmet_edf <- rbind(cmet_edf,
                        cbind(dsName=paste0("1_",i), 
                            transform="classify", phase="training",
                            jentropies(dFvector[inTrain,],
                                      hatKdataset[inTrain,])),
                        cbind(dsName=paste0("1_",i),  
                            transform="classify", phase="testing",
                            jentropies(dFvector[-inTrain,],
                                      hatKdataset[-inTrain,]))
                        )
            smet_edf <- rbind(smet_edf, 
                        cbind(dsName=paste0("1_",i), 
                              group="prediction", phase="training",
                              sentropies(hatKdataset[inTrain,],type="dual")),
                        cbind(dsName=paste0("1_",i), 
                              group="prediction", phase="testing",
                              sentropies(hatKdataset[-inTrain,],type="dual"))
                        )
        }# End going over all of the selected features
}
if (inDebugMode){
    # show the split entropies
    str(cmet_edf)
    #show all entropies for a data base
    print(filter(smet_edf, dsName == "Ionosphere"))
    print(filter(cmet_edf, dsName == "Ionosphere"))
}
```

Creating SMETS and CMETS for every possible database.

```{r}
thisPhase <- "testing"
#dsname <- "Ionosphere"
smets <- list()
#for(dsname in dsNames){
    setdf <- smet_edf %>%  
        filter(phase == "testing") 
    smet <- ggmetern(setdf, fancy) +
        geom_point(aes(color=group,shape=group), size=5) #+
        #scale_shape_manual(values=c(4, 20, 1)) +
        #labs(color="grouping")# +
#}
if (writePlots){
    #dev.off()#Necessary to do the textual plot.
    ggsave(str_interp("ID_ICA_smet_${dsName}.jpeg"), 
           plot=smet)
} else
    smet
```

```{r}
cmets <- list()
cetdf <- cmet_edf %>% 
    filter(type != "XY" & phase==thisPhase) 
cmet <- ggmetern(cetdf, fancy) +
    geom_point(aes(shape=type,color=transform),size=5) +
    scale_shape_manual(values=c(4, 20, 1)) #+
#labs(color="transfo name", shape="Var type")# +
if (writePlots){
    #dev.off()#Necessary to do the textual plot.
    ggsave(str_interp("ID_ICA_cmet_${dsName}.jpeg"), 
           plot=cmet)
} else
    cmet
```

# Analysis of performance

```{r}
# The colors for the different feature sets for the plots. 
mncols <- min(10,ncols)
if (fancy){
    orderingColors <- rev(terrain.colors(mncols+1))[1:mncols + 1]
} else {
    orderingColors <- rev(gray(0:mncols / mncols))[1:mncols + 1]
    #orderingColors <- scale_colour_grey(end=0.9)#Not for discrete levels
}
# The shapes for the different types of transform
#transformShapes <- c("log"=4, "PCA"=1, "ICA"=5)#no fill, void
#transformShapes <- c("log"=4, "PCA"=20, "ICA"=18)#no fill, solid
transformShapes <- c("log"=4, "PCA"=21, "ICA"=23)#no fill, solid
sourceShapes <- c("none"=8, "log"=4, "PCA"=21, "ICA"=23)#no fill, solid
typeShapes <- c("X" = 4, "Y" = 1, "ALL" = 10, "XY" = 20)

e2e_cmet <-  tibble()# Gathers the end-to-end entropies
for( i in (1:(min(10, ncol(ds))))){
    e2e_cmet <- rbind(e2e_cmet,
                      cbind(
                          dsName=paste0("1_",i), 
                          phase="training",
                          jentropies(trainCM[[i]]$table)),
                      cbind(
                          dsName=paste0("1_",i), 
                          phase="testing",
                          jentropies(testCM[[i]]$table))
                      )
}
e2ecmet <- ggmetern(e2e_cmet %>% filter(type=="XY"), fancy) +
    geom_point(aes(shape=phase,color=dsName),size=5) +
    scale_shape_manual(values=c(4, 20, 1)) +
    scale_colour_manual(values=orderingColors)
    #labs(color="transfo name", shape="Var type")# +
if (writePlots){
    #dev.off()#Necessary to do the textual plot.
    ggsave(str_interp("ID_ICA_e2e_cmet_${dsName}.jpeg"), 
           plot=e2ecmet)
} else
    e2ecmet
```
```



# Postscriptum

```{r ps}
sessionInfo()
```
