---
title: "Multilabelling Evaluation with the CMET"
author: "Francisco J. Valverde-Albacete & Carmen Peláez Moreno"
date: "5/15/2017"
output: html_document
---

# Introduction

This vignette esplores the capabilities of the Channel Multivariate Entropy Triangle (CMET) to explore the results of multilabel classification.

This vignette follows the scheme to carry out multilabel evaluation proposed by 

https://mlr-org.github.io/mlr-tutorial/devel/html/multilabel/index.html

# Environment construction

## Knitting options

```{r setup, include=FALSE}
#knitr::opts_chunk$set(dev = 'pdf') # plots in pdf, better for publication
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment=NA, fig.width=6, fig.height=6)
#knitr::opts_chunk$set(comment=NA, fig.width=6, fig.height=4)
knitr::opts_chunk$set(warning=FALSE)# Should not appear in the knitted document
```

## Library loading

```{r libraries, include=FALSE}
library(mlr)
library(tidyverse)  # Acceding to that infamous Mr. Wickham's requests!
# library(dplyr)     # That infamous Mr. Wickham!
# library(tidyr)     # Tidying tall & wide dataframes
library(infotheo)  # The functionality provided by this has to be rerouted through entropies
library(entropies) # This package
library(ggtern)    # Excellent package for ternary diagrams in the gg tradition
library(vcd)       # Categorical benchmarks
library(mlbench)   # ml benchmarkss
```

# Multilabel Classification

## Data preparation

```{r data}
yeast = getTaskData(yeast.task)
labels = colnames(yeast)[1:14]
yeast.task = makeMultilabelTask(id = "multi", data = yeast, target = labels)
yeast.task
```

### Source exploration with the SMET

```{r source_exploration}
sedf <- frame()
labels <- yeast[,1:14]
sedf <- rbind(
        sedf,
        cbind(dsName = "yeast", sentropies(labels))
        # getDatasetSourceEntropies(yeast.task, 
        #     #dsName, 
        #     #className = dsRecord$className,
        #     #withClass=TRUE, 
        #     type="total")
        )
sedf
```
We visualize the real labels with the Source Multivariate Entropy Triangle (SMET)

```{r SMET_vis}
smet <-  ggmetern(sedf, fancy) + 
    geom_point(mapping=aes(colour=name), size=3)+
    scale_shape_manual(values=1:ncol(labels)) #+
    #labs(shape="Dataset")
if (fancy){
    smet <- smet + ggtitle("Source Multivariate Entropy for labels")
}
smet
#ggsave("disgregated_labels_label.jpeg", plot=smet)

```

## Classifier training

### Adaptation methods

```{r adaptation}
# lrn.rfsrc = makeLearner("multilabel.randomForestSRC", predict.type="response")
# lrn.rfsrc
lrn.rFerns = makeLearner("multilabel.rFerns")
lrn.rFerns
```

### Problem transformation methods

```{r transformation}
lrn.br = makeLearner("classif.rpart", predict.type = "prob")
lrn.br = makeMultilabelBinaryRelevanceWrapper(lrn.br)
lrn.br
lrn.br2 = makeMultilabelBinaryRelevanceWrapper("classif.rpart")
lrn.br2

```

### Training

```{r training}
mod = train(lrn.br, yeast.task)
mod = train(lrn.br, yeast.task, subset = 1:1500, weights = rep(1/1500, 1500))
mod

mod2 = train(lrn.rFerns, yeast.task, subset = 1:100)
mod2
```

### Prediction

```{r prediction}

pred = predict(mod, task = yeast.task, subset = 1:10)
pred = predict(mod, newdata = yeast[1501:1600,])
names(as.data.frame(pred))


pred2 = predict(mod2, task = yeast.task)
names(as.data.frame(pred2))
```

### Performance measurement

```{r performance}
listMeasures("multilabel")
getDefaultMeasure(yeast.task)
#performance(pred)
performance(pred, measures = list(#multilabel.subset01,
                                  hamloss, 
                                  #multilabel.acc,
                                  #multilabel.f1, 
                                  timepredict)
            )
#performance(pred2)
performance(pred2, measures =list(#multilabel.subset01,
                                  hamloss, 
                                  #multilabel.acc,
                                  #multilabel.f1, 
                                  timepredict)
            )
```

# Evaluation on the CMET

```{r evaluation_br}
library(dplyr)
edf <- data.frame()
# the predictions are frames
br.truth <- pred$data[,1:14]
br.predicted <- pred$data[,29:42]
#summary(br.predicted)
assertthat::are_equal(ncol(br.truth), 
                      ncol(br.predicted))
edf <- rbind(edf,
             cbind(method="br",
                   jentropies(br.truth,br.predicted)
                   )
             )
```

```{r evaluation_rFerns}
edf <- rbind(edf,
             cbind(method="rFerns",
                   jentropies(pred2$data[,1:14], pred2$data[,15:28])
                   )
             )
edf
```

#### Visualization

Visualize both the aggregate and the split entropies.

```{r, fig.width=12, fig.height=16}
fancy <- TRUE
#fancy <- FALSE
#select some entropies to visualize
cmet <- ggmetern(edf %>% filter(type != "XY"), 
                 #filter(edf, type != "XY"),#alternative not to visualize the aggregate
                 fancy) +
    geom_point(aes(color=type, shape=method), size=3) +
    scale_shape_manual(values=c(4, 20, 1)) #+
    #labs(color="Dataset name", shape="Var type")
cmet #+ facet_wrap(~dsName, ncol=2)
#dev.off()#Necessary to do the textual plot.
#ggsave(str_interp("entropyEndToEnd_${dsName}_epoch_${thisEpoch}.jpeg"), plot=e2ePlot)
```

Interpretation...

## TODO: interpret these results!!!



# Postscriptum

```{r ps}
sessionInfo()
```


